---
title: "Lung Cancer Risk Analysis"
author: "Winn Wu"
date: "`r Sys.Date()`"
output: html_document
---

```{r, r-chunk settings, echo = FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    warning = FALSE,
    message = FALSE,
    error = FALSE,
    cache = FALSE
)
```

```{r, libraries}
library(ggplot2)
library(sl3)
library(tidyverse)
library(gtsummary)
library(scales)
library(gt)
```


```{r, data loading and preprocessing}
data <- read.csv("lung_cancer_dataset.csv")

# factorize the categorical variables
data$gender <- factor(data$gender)
data$radon_exposure <- factor(data$radon_exposure)
data$asbestos_exposure <- factor(data$asbestos_exposure)
data$secondhand_smoke_exposure <- factor(data$secondhand_smoke_exposure)
data$copd_diagnosis <- factor(data$copd_diagnosis)
data$alcohol_consumption <- factor(data$alcohol_consumption)
data$family_history <- factor(data$family_history)
data$lung_cancer <- factor(data$lung_cancer)
```

### Introduction
This dataset is from Kaggle, designed for lung cancer risk analysis. It contains 50,000 clean, processed profiles from patients who constantly smoke cigarettes, with features based on established lung cancer risk factors such as smoking exposure, environmental exposures, and chronic lung conditions. All data is synthetic and designed to reflect realistic risk factor distributions while maintaining patient privacy.

#### Key Variables
- `age`: Age of the patient (years)
- `gender`: Gender of the patient (Male, Female)
- `pack_years`: Smoking exposure (years * packs/day)
- `radon_exposure`: Exposure to radon (Low, Medium, High)
- `asbestos_exposure`: Exposure to asbestos (No, Yes)
- `secondhand_smoke_exposure`: Exposure to secondhand smoke (No, Yes)
- `copd_diagnosis`: Diagnosis of chronic obstructive pulmonary disease (No, Yes)
- `alcohol_consumption`: Alcohol consumption (None, Moderate, Heavy)
- `family_history`: Family history of lung cancer (No, Yes)
- `lung_cancer`: Diagnosis of lung cancer (No, Yes)

### Data Analysis Objectives
In this project, we will analyze the data to present some basic descriptive statistics and answer the following question:

For the population represented by this dataset, what is the effect of smoking exposure on the risk of developing lung cancer?



\newpage

### Descriptive Statistics

The following table summarizes the characteristics of all variables in the dataset:

\FloatBarrier

```{r, descriptive statistics, results='asis'}
# get a table one for the data, excluding ID and showing all levels
tbl_summary_object <- data %>%
  select(-patient_id) %>%  # exclude the id variable
  tbl_summary(
    type = list(
      # Force binary variables to show both levels
      radon_exposure ~ "categorical",
      asbestos_exposure ~ "categorical", 
      secondhand_smoke_exposure ~ "categorical",
      copd_diagnosis ~ "categorical",
      alcohol_consumption ~ "categorical",
      family_history ~ "categorical",
      lung_cancer ~ "categorical",
      gender ~ "categorical"
    ),
    statistic = list(
      # For categorical variables, show n (%)
      all_categorical() ~ "{n} ({p}%)",
      # For continuous variables, show median (IQR)
      all_continuous() ~ "{median} ({p25}, {p75})"
    )
  ) %>%
  # Add variable labels for better readability
  modify_header(label ~ "**Variable**") %>%
  modify_spanning_header(c("stat_0") ~ "**Overall, N = {N}**")

# Convert to gt with larger formatting
tbl_summary_object %>% 
  as_gt() %>%
  tab_options(
    table.font.size = 12,
    data_row.padding = px(8),
    summary_row.padding = px(8),
    grand_summary_row.padding = px(8),
    footnotes.font.size = 10,
    source_notes.font.size = 10,
    heading.subtitle.font.size = 13,
    heading.title.font.size = 15,
    table.width = pct(100),
    column_labels.font.weight = "bold",
    table.margin.left = px(10),
    table.margin.right = px(10)
  )

```

\FloatBarrier

\vspace{1cm}

This table provides a comprehensive overview of the dataset, showing the distribution of each variable across all patients. For continuous variables, the table shows the median and interquartile range, and shows the number of observations and percentage for categorical variables in each level.

It is useful for understanding the baseline characteristics of the population and for identifying any potential patterns or differences in the data.


### Data Analysis
```{r, data analysis, fig.height=4, fig.width=7}
# define a learner to run the classification task
lr <- Lrnr_gam$new()

# define the covariates
w_vars <- c(
    "age", "gender", "radon_exposure", "asbestos_exposure", "pack_years",
    "secondhand_smoke_exposure", "copd_diagnosis", "alcohol_consumption",
    "family_history"
)

# define the outcome
y_var <- c("lung_cancer")

# define a task
task <- make_sl3_Task(
    data = data,
    outcome = y_var,
    covariates = w_vars
)

# fit the task
fit <- lr$train(task)

# create a dummy data for prediction with proper factor levels
dummy_data <- data.frame(
    age = 50,
    gender = factor("Male", levels = levels(data$gender)),
    radon_exposure = factor("Low", levels = levels(data$radon_exposure)),
    asbestos_exposure = factor("No", levels = levels(data$asbestos_exposure)),
    pack_years = seq(0, 50, 0.1),
    secondhand_smoke_exposure = factor("No", levels = levels(data$secondhand_smoke_exposure)),
    copd_diagnosis = factor("No", levels = levels(data$copd_diagnosis)),
    alcohol_consumption = factor("None", levels = levels(data$alcohol_consumption)),
    family_history = factor("No", levels = levels(data$family_history))
)

# define the new task for prediction (no outcome variable needed)
new_task <- make_sl3_Task(
    data = dummy_data,
    covariates = w_vars
)

# predict the probability of developing lung cancer
pred <- fit$predict(new_task)

# plot the result
ggplot(data = dummy_data, aes(x = pack_years, y = pred)) +
    geom_line(color = "darkred", size = 1.2) +
    scale_y_continuous(
        labels = scales::percent_format(accuracy = 1),
        limits = c(0, max(pred) * 1.05),
        breaks = scales::pretty_breaks(n = 6)
    ) +
    scale_x_continuous(
        breaks = scales::pretty_breaks(n = 8),
        limits = c(0, 50)
    ) +
    labs(
        title = "Predicted Lung Cancer Risk by Smoking Exposure",
        subtitle = "50-year-old male with no other risk factors",
        x = "Pack Years (Packs per day × Years of smoking)",
        y = "Predicted Probability of Lung Cancer",
        caption = "Based on GAM model fitted to lung cancer risk dataset"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray60"),
        plot.caption = element_text(size = 9, color = "gray50"),
        axis.title = element_text(size = 11, face = "bold"),
        axis.text = element_text(size = 10),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "gray90", size = 0.5),
        plot.background = element_rect(fill = "white", color = NA)
    )

```

\vspace{1cm}

This plot shows the predicted probability of developing lung cancer of a 50-year-old male with no other risk factors by smoking exposure. The plot is based on a GAM model fitted to the lung cancer risk dataset. The plot shows that the risk of developing lung cancer increases with smoking exposure when smoking exposure is lower than about 32 pack years, and supurisingly decreases when smoking exposure is higher than about 32 pack years. Probabaly because there is some confounding effect that is not accounted for in the model.